---
title: "04_risk_score"
output: html_document
date: "2026-05-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
# If rms or dcurves are not yet in your renv, install them first:
#   renv::install("rms"); renv::install("dcurves"); renv::snapshot()

library(tidyverse)
library(here)
library(DT)
library(rms)       # Harrell's regression modelling strategies: lrm, validate, calibrate
library(dcurves)   # decision curve analysis
library(broom)     # tidy() for clean coefficient tables
```


## 1. Analytic sample

Reproduce the Phase 3 analytic dataset so this script is fully self-contained.
Criteria: baseline exam (PERIOD == 1), no prevalent CHD at baseline (PREVCHD == 0),
complete cases on all seven predictors.

```{r load_data, message=FALSE}
framingham <- read_csv(here("data/raw/framingham.csv"))
```

```{r prep}
analytic <- framingham %>%
  filter(PERIOD == 1, PREVCHD == 0) %>%
  select(RANDID, CVD, AGE, SYSBP, TOTCHOL, BMI, GLUCOSE, CURSMOKE, DIABETES) %>%
  drop_na()

analytic %>%
  summarise(n = n(), n_cvd = sum(CVD), cvd_rate = round(mean(CVD), 3))
```


## 2. Fit the logistic model (two objects, same formula)

We need two fit objects:

- `cvd_glm` (base R `glm`) — used for coefficient extraction and point-score derivation.
  Familiar from Phase 3; coefficients are raw log-odds.
- `cvd_lrm` (rms `lrm`) — the same model fit through the rms framework. The rms object
  stores the design matrix and outcome vector internally, which is what enables the
  bootstrap routines `validate()` and `calibrate()` later.

Both produce identical predicted probabilities; the difference is purely infrastructure.

```{r glm_fit}
cvd_glm <- glm(
  CVD ~ AGE + SYSBP + TOTCHOL + BMI + GLUCOSE + CURSMOKE + DIABETES,
  family = binomial,
  data   = analytic
)

tidy(cvd_glm, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  datatable(caption = "Phase 3 model — odds ratios (recap)")
```

```{r lrm_fit}
# rms requires a datadist object: a summary of each predictor's distribution.
# lrm() uses this to compute reference-cell predictions and "typical patient" summaries.
dd <- datadist(analytic)
options(datadist = "dd")

cvd_lrm <- lrm(
  CVD ~ AGE + SYSBP + TOTCHOL + BMI + GLUCOSE + CURSMOKE + DIABETES,
  data = analytic,
  x    = TRUE,   # store design matrix — required for validate() / calibrate()
  y    = TRUE    # store outcome vector — same reason
)

cvd_lrm
```

The `lrm` output reports Dxy (Somers' D concordance index) and C (the AUC). These should
match the pROC AUC from Phase 3 to within rounding — a good sanity check.


## 3. Convert to a point-based risk score

### Why bother with a point score?

A logistic regression model produces an individualized predicted probability, but it
requires a computer to evaluate. In clinical settings — especially resource-limited ones,
or for quick triage — a simple integer score that a clinician can compute on paper has
real utility. This is how the original Framingham Risk Score (Wilson et al., 1998) and
the CHADS₂ stroke-risk score work.

The key insight is that a logistic model's **linear predictor** (LP = β₀ + Σ βⱼ xⱼ) is
a weighted sum of the predictors. A point score is just a scaled, rounded version of that
same sum — it preserves the ordering of predicted risk almost perfectly.

### Sullivan scoring method (Sullivan et al., 2004)

For each predictor j, assign integer points as:

    points_j = round( βⱼ × incrementⱼ / β_ref )

where:
- βⱼ  is the regression coefficient (log-odds scale)
- incrementⱼ  is a clinically meaningful step size (e.g. 10 mmHg of SBP)
- β_ref  is a common scaling denominator — we use β_AGE (per 1 year of age)

Dividing by β_ref anchors the scale: 1 point of score = the same risk increase as
1 additional year of age. Everything else is expressed relative to that anchor.

```{r point_score_derivation}
coefs    <- coef(cvd_glm)
beta_ref <- coefs["AGE"]   # anchor: 1 year of age

# Clinically meaningful increments — chosen to produce legible integers
increments <- c(
  AGE      = 1,    # 1 year
  SYSBP    = 10,   # 10 mmHg
  TOTCHOL  = 40,   # 40 mg/dL (~1 mmol/L conversion step)
  BMI      = 5,    # 5 kg/m²
  GLUCOSE  = 20,   # 20 mg/dL
  CURSMOKE = 1,    # binary: 1 = smoker
  DIABETES = 1     # binary: 1 = diabetic
)

score_table <- tibble(
  predictor  = names(increments),
  beta       = coefs[names(increments)],
  increment  = increments,
  points_raw = beta * increment / beta_ref,
  points     = round(points_raw)
)

score_table %>%
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  datatable(caption = "Point score derivation — Sullivan method (β_ref = β_AGE)")
```

Read the `points` column as: *+1 point for each additional year of age; +N points for
each increment of the continuous predictors; smokers and diabetics receive a fixed bonus.*

### Apply score to the analytic sample

Instead of rounding each predictor's contribution individually (which accumulates rounding
error), we compute the exact scaled linear predictor and round the total once at the end.
This is mathematically equivalent to summing the per-predictor points but more precise.

```{r apply_score}
intercept <- coefs["(Intercept)"]

analytic_scored <- analytic %>%
  mutate(
    # Scaled LP: same as the logistic LP but divided by β_ref so the unit is "age-years"
    lp_scaled  = (
      AGE      * coefs["AGE"]      +
      SYSBP    * coefs["SYSBP"]    +
      TOTCHOL  * coefs["TOTCHOL"]  +
      BMI      * coefs["BMI"]      +
      GLUCOSE  * coefs["GLUCOSE"]  +
      CURSMOKE * coefs["CURSMOKE"] +
      DIABETES * coefs["DIABETES"]
    ) / beta_ref,
    risk_score = round(lp_scaled)
  )

# Distribution of scores by CVD status
analytic_scored %>%
  ggplot(aes(x = risk_score, fill = factor(CVD))) +
  geom_histogram(binwidth = 2, position = "identity", alpha = 0.6, colour = "white") +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "tomato"),
    labels = c("0" = "No CVD",    "1" = "CVD"),
    name   = "Outcome"
  ) +
  labs(
    x     = "Risk score (integer points)",
    y     = "Count",
    title = "Distribution of integer risk scores by CVD status"
  ) +
  theme_minimal()
```

### Score-to-probability lookup table

The relationship between the integer score and predicted probability follows directly from
logistic regression:

    P(CVD) = 1 / ( 1 + exp( –(β₀ + score × β_ref) ) )

This is the lookup table you would print on a pocket card or clinical decision-support tool.

```{r score_to_prob}
score_prob_table <- tibble(
  score = seq(
    floor(min(analytic_scored$risk_score)),
    ceiling(max(analytic_scored$risk_score)),
    by = 5
  )
) %>%
  mutate(
    lp            = intercept + score * beta_ref,
    predicted_prob = round(1 / (1 + exp(-lp)), 3)
  )

datatable(
  score_prob_table,
  caption = "Score-to-probability lookup table (every 5 points)"
)
```


## 4. Bootstrap internal validation — Harrell framework

### The overfitting problem

When we fit a logistic model on a dataset, the parameter estimates are tuned — even
slightly, even with many observations — to the noise in that particular sample. If we
then evaluate the AUC by asking the model to predict the very outcomes it was optimised
on, we get an artificially inflated estimate. That gap between apparent and true
population performance is called **optimism**.

Think of it like grading a student on the same exam they studied from: you'll
systematically overestimate how much they actually understand.

The degree of optimism depends on:
- Sample size (smaller N → more optimism)
- Number of predictors relative to events (fewer events per predictor → more optimism)
- Model complexity

With ~940 events and 7 predictors (~134 EPV, events-per-variable), we have a relatively
generous dataset. But internal validation is good practice regardless — and it's
explicitly in the TRIPOD reporting guideline for clinical prediction models.

### The Harrell bootstrap procedure

Harrell's method (Efron 1983, Harrell 1996) estimates optimism without splitting the data:

1. Record the **apparent** metric (e.g. AUC) on the full training set.
2. Repeat B times:
   a. Draw a bootstrap sample of n rows **with replacement** from the training data.
   b. Fit the full model specification on the bootstrap sample → evaluate AUC on that
      same bootstrap sample. This step mimics "apparent" AUC in a hypothetical new dataset.
   c. Apply the bootstrap-fit model to the **original** data → compute AUC again.
      This step mimics "new patient" performance using the bootstrap model as a proxy.
   d. Optimism_b = AUC_bootstrap_sample – AUC_original_data
3. Mean optimism = average of Optimism_b across all B iterations.
4. Corrected AUC = Apparent AUC – Mean optimism.

The beauty of this approach: you never discard training data (unlike a train/test split),
and the estimate of optimism is stable across random seeds given sufficient B.

```{r bootstrap_validation}
set.seed(42)
val <- validate(cvd_lrm, B = 200)
val
```

```{r auc_summary}
# Somers' Dxy relates to AUC: Dxy = 2 × (AUC – 0.5), so AUC = Dxy/2 + 0.5
apparent_auc  <- (val["Dxy", "index.orig"] / 2) + 0.5
optimism      <-  val["Dxy", "optimism"]   / 2
corrected_auc <- apparent_auc - optimism

tibble(
  metric = c("Apparent AUC (training set)",
             "Mean optimism (B = 200 bootstrap)",
             "Optimism-corrected AUC"),
  value  = round(c(apparent_auc, optimism, corrected_auc), 3)
) %>%
  datatable(caption = "Bootstrap validation — AUC summary")
```

Interpret the corrected AUC as your best honest estimate of how the model would
discriminate in a fresh sample of similar patients — without having to actually collect
new data.


## 5. Calibration slope and intercept

Discrimination (AUC) tells you whether the model separates events from non-events.
**Calibration** asks a different question: are the predicted probabilities numerically
accurate? A model can discriminate perfectly but still tell every patient their risk is
twice what it actually is.

Two summary statistics capture this:

- **Calibration intercept (calibration-in-the-large):** the mean log-odds of the
  predicted probabilities versus the observed event rate. Should be 0. A positive value
  means the model under-predicts on average; negative means over-prediction.

- **Calibration slope:** the regression coefficient when you regress observed outcomes on
  the model's linear predictor in a new dataset. Should be 1.0. A slope < 1 is the
  signature of overfitting — the model spreads predicted probabilities too wide; extreme
  predictions (very high and very low) are shrunk toward the mean in reality.

Bootstrap-corrected calibration slope is one of the key summary statistics in TRIPOD
reporting. Values above ~0.90 are generally considered acceptable for internal validation
on a reasonably-sized dataset.

```{r calibration, warning=FALSE}
set.seed(42)
cal <- calibrate(cvd_lrm, B = 200)
plot(
  cal,
  main = "Bootstrap-corrected calibration — CVD logistic model",
  xlab = "Predicted probability",
  ylab = "Observed event rate"
)
```

```{r calibration_slope}
# The validate() output includes the calibration slope directly
cal_slope     <- val["Slope", "index.corrected"]
cal_intercept <- val["Intercept", "index.corrected"]

tibble(
  statistic = c("Calibration intercept (corrected)", "Calibration slope (corrected)"),
  value     = round(c(cal_intercept, cal_slope), 3),
  ideal     = c(0, 1)
) %>%
  datatable(caption = "Calibration summary statistics — bootstrap corrected")
```


## 6. Decision curve analysis

### Beyond ROC: making decisions under uncertainty

ROC curves and AUC answer "how well does the model discriminate?" but they do not answer
the question a clinician actually faces: **"At what probability threshold should I treat,
and does using this model to make that decision actually help patients compared to my
alternatives?"**

Decision curve analysis (Vickers & Elkin, 2006) addresses this directly by computing
**net benefit** across a range of threshold probabilities.

The threshold probability (pt) represents the clinician's implicit cost ratio: the
probability at which they are indifferent between treating and not treating. A threshold
of 20% means: "I'd rather treat 4 people who don't need it to prevent 1 missed event."

Net benefit is defined as:

    NB = (True Positives / N) – (False Positives / N) × pt / (1 – pt)

The penalty term (pt / (1 – pt)) scales up the cost of false positives according to the
threshold. At low thresholds (e.g. 10%), false positives are cheap — the treat-all
strategy looks good. At high thresholds, they are expensive — selective treatment wins.

Three curves are always plotted for comparison:
1. **Model:** NB using the predicted probabilities to select who to treat.
2. **Treat all:** assume every patient has the event → NB = event rate – (1 – event rate) × pt/(1–pt).
3. **Treat none:** NB = 0 by definition.

The model is clinically useful at threshold pt if its NB curve is above both treat-all
and treat-none.

```{r dca}
# dcurves requires predicted probabilities as a column in the data frame
analytic_dca <- analytic %>%
  mutate(pred_prob = predict(cvd_glm, type = "response"))

dca_result <- dca(
  CVD ~ pred_prob,
  data       = analytic_dca,
  label      = list(pred_prob = "Logistic model"),
  thresholds = seq(0.05, 0.50, by = 0.01)
)

plot(dca_result, smooth = TRUE) +
  labs(
    title    = "Decision curve analysis — CVD logistic model",
    subtitle = "Net benefit across threshold probabilities (5% – 50%)",
    x        = "Threshold probability",
    y        = "Net benefit"
  ) +
  theme_minimal()
```

Read the DCA plot as: for any threshold probability on the x-axis, the y-axis tells you
the net benefit per 100 patients. If the model's curve is above the "treat all" line, the
model adds value over blanket treatment at that threshold; if above "treat none," it adds
value over watchful waiting.


## Summary

| Component | Method | Key output |
|---|---|---|
| Point score | Sullivan scaling (β / β_ref) | Integer score + lookup table |
| Internal validation | Harrell bootstrap (B=200) | Optimism-corrected AUC |
| Calibration | Bootstrap-corrected slope/intercept | Slope ≈ 1, intercept ≈ 0 |
| Clinical utility | Decision curve analysis | Net benefit vs treat-all/none |

**TRIPOD checklist items covered here:** model performance (discrimination + calibration),
internal validation method, presentation of absolute predicted probabilities.
