---
title: "05_survival"
output: html_document
date: "2026-05-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
# New package: renv::install("survminer"); renv::snapshot()
library(tidyverse)
library(here)
library(DT)
library(survival)    # core: Surv(), survfit(), survdiff(), coxph(), cox.zph()
library(survminer)   # ggplot2 KM plots: ggsurvplot()
library(broom)       # tidy() for clean Cox output
```


## 1. Analytic sample

```{r}
# read in and clean df

framingham <- read_csv(here("data", "raw", "framingham.csv"))

analytic <- framingham %>% 
  filter(PERIOD == 1, PREVCHD == 0) %>% 
  select(
    RANDID, CVD, AGE, SYSBP, TOTCHOL, BMI, GLUCOSE, CURSMOKE, DIABETES, SEX, TIMECVD) %>% 
  drop_na() %>% 
  mutate(timecvd_yr = TIMECVD / 365.25) %>% 
  select(-TIMECVD)

# verify df

analytic %>%
  summarise(
    n          = n(),
    n_cvd      = sum(CVD),
    cvd_rate   = round(mean(CVD), 3)
  )
```



## 2. The survival object

# Before any modelling or plotting, we package the outcome into a Surv object.
# Surv(time, event) pairs each patient's follow-up time with their event status.
# Censored observations (CVD == 0) contribute their survival time as a lower bound.

```{r surv_object}
surv_obj <- Surv(time  = analytic$timecvd_yr,
                 event = analytic$CVD)

# Quick look: the + symbol marks censored observations
head(surv_obj, 20)
```


## 3. Kaplan-Meier — overall survival curve

```{r km_overall}
# survfit() estimates the survival function from a Surv object.
# The formula ~ 1 means "one curve for the whole sample, no grouping."
km_overall <- survfit(surv_obj ~ 1, data = analytic)

summary(km_overall, times = c(2, 5, 8, 10, 12))
```

```{r km_overall_plot}
ggsurvplot(
  km_overall,
  data       = analytic,
  conf.int   = TRUE,
  risk.table = TRUE,        # patients still at risk shown below the curve
  xlab       = "Time (years)",
  ylab       = "CVD-free probability",
  title      = "Kaplan-Meier — overall CVD-free survival",
  ggtheme    = theme_minimal()
)
```


## 4. KM stratified by key predictors

```{r km_smoking}
km_smoke <- survfit(surv_obj ~ CURSMOKE, data = analytic)

ggsurvplot(
  km_smoke,
  data       = analytic,
  conf.int   = TRUE,
  risk.table = TRUE,
  legend.labs = c("Non-smoker", "Smoker"),
  xlab       = "Time (years)",
  ylab       = "CVD-free probability",
  title      = "KM by smoking status",
  ggtheme    = theme_minimal()
)
```

```{r km_diabetes}
km_diab <- survfit(surv_obj ~ DIABETES, data = analytic)

ggsurvplot(
  km_diab,
  data        = analytic,
  conf.int    = TRUE,
  risk.table  = TRUE,
  legend.labs = c("No diabetes", "Diabetes"),
  xlab        = "Time (years)",
  ylab        = "CVD-free probability",
  title       = "KM by diabetes status",
  ggtheme     = theme_minimal()
)
```

```{r km_sex}
km_sex <- survfit(surv_obj ~ SEX, data = analytic)

ggsurvplot(
  km_sex,
  data        = analytic,
  conf.int    = TRUE,
  risk.table  = TRUE,
  legend.labs = c("Male", "Female"),
  xlab        = "Time (years)",
  ylab        = "CVD-free probability",
  title       = "KM by sex",
  ggtheme     = theme_minimal()
)
```


## 5. Log-rank test

# The log-rank test asks: are these survival curves significantly different,
# or could the gap be explained by chance? One test per stratifier.

```{r logrank}
logrank_smoke <- survdiff(surv_obj ~ CURSMOKE, data = analytic)
logrank_diab  <- survdiff(surv_obj ~ DIABETES,  data = analytic)
logrank_sex   <- survdiff(surv_obj ~ SEX,        data = analytic)

logrank_smoke
logrank_diab
logrank_sex
```


## 6. Cox proportional hazards model

# Same seven predictors as the Phase 3 logistic model — allows direct comparison
# of hazard ratios (HR) to the odds ratios (OR) we already computed.

```{r cox_fit}
cox_fit <- coxph(
  surv_obj ~ AGE + SYSBP + TOTCHOL + BMI + GLUCOSE + CURSMOKE + DIABETES,
  data = analytic
)

summary(cox_fit)
```

```{r cox_tidy}
tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  datatable(caption = "Cox model — hazard ratios with 95% CI")
```


## 7. Proportional hazards assumption check

# The Cox model assumes that the hazard ratio for each predictor is constant
# over time — the "proportional hazards" assumption.
# cox.zph() tests this using scaled Schoenfeld residuals.
# A significant p-value (< 0.05) flags a predictor whose effect varies over time.

```{r ph_test}
ph_test <- cox.zph(cox_fit)
ph_test
```

```{r ph_plot}
# Plot scaled Schoenfeld residuals for each predictor.
# A flat, horizontal line supports the PH assumption.
# A sloped or curved line suggests the hazard ratio changes over time.
ggcoxzph(ph_test)
```


## 8. HR vs OR comparison — what does survival analysis add?

```{r hr_or_compare}
# Side-by-side: hazard ratios from Cox vs odds ratios from Phase 3 logistic
cox_hr <- tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high) %>%
  rename(HR = estimate, HR_lo = conf.low, HR_hi = conf.high)

# Re-fit Phase 3 logistic model for direct comparison
glm_fit <- glm(
  CVD ~ AGE + SYSBP + TOTCHOL + BMI + GLUCOSE + CURSMOKE + DIABETES,
  family = binomial, data = analytic
)

glm_or <- tidy(glm_fit, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  select(term, estimate, conf.low, conf.high) %>%
  rename(OR = estimate, OR_lo = conf.low, OR_hi = conf.high)

left_join(cox_hr, glm_or, by = "term") %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  datatable(caption = "HR (Cox) vs OR (logistic) — same predictors, same sample")
```
